Ionospheric Disturbances in Mexican Territory Produced by Objects Entering the Atmosphere from Space \tnoteref{tnote1}

### 2022-11-29 paper/appendix.tex
Update ELSEVIER paper (which compiles) to overleaf at 28-November-2022

% This is tha appendix, where we may insert detailed mathematical proccedures or other data which may be useful but also will make the main
% text too large so this should be consulted only in case of neccesity

\section{Estimation of distances to meteors}

The distance between the GOES satellite and the meteor, neccesary to estimate the radiated energy is calculated as follows:

\begin{align}
  R = \left|\vec{r}_{GLM} - \vec{r}_{obj} \right|
\end{align}

where $\vec{r}_{GLM}$ is the vector position of the GLM satellite and $\vec{r}_{obj}$ is the vector position of the meteor.

In cartesian coordinates, the distance $R$ is given by:

\begin{align}
  R &= \left((x_{GLM}-x_{obj})^2 + (y_{GLM}-y_{obj})^2 + (z_{GLM}-z_{obj})^2\right)^{1/2} \label{eq:R}\\
\end{align}

The transformation to spherical coordinates is given by:

\begin{align}
  x = r\cos\phi\cos\theta \label{eq:xtoR}\\
  y = r\sin\phi\cos\theta \label{eq:ytoR}\\
  z = r\sin\theta \label{eq:ztoR}
\end{align}

Where $r$ is measured from the center of the earth, $ -180^\circ < phi <= 180^\circ$ represents the longitude; is positive at east of Greenwich meridian, and negative eastwards. $\-90^\circ <= \theta <= 90^\circ$ represents the latitude and is positive at the north of equator and negative southwards.

Substituting the transform (\ref{eq:xtoR} - \ref{eq:ztoR}) into (\ref{eq:R}), using elemental trigonometry and considering both GLM satellites lie into the equator ($\theta_{GLM}=0$) we get:

\begin{align}
  R^2 &= r_{GLM}^2 + r_{obj}^2 - 2r_{GLM}r_{obj}f(\theta_{obj}, \phi_{obj}, \phi_{GLM}) \label{eq:R2}\\
  \mathrm{where:} & f(\theta_{obj}, \phi_{obj}, \phi_{GLM}) = \cos\theta_{obj}\cos\left(\phi_{GLM}-\phi_{obj}\right)
\end{align}

Since $r_{GLM}$ and $r_{obj}$ are measured from the center of the arth ge find that:

\begin{align}
  r_{GLM} &= r_{earth} + h_{GLM} \label{eq:rg}\\
  r_{obj} &= r_{earth} + h_{obj} \label{eq:ro}
\end{align}

Substituting (\ref{eq:rg}, \ref{eq:ro}) into \ref{eq:R2} and considering that $h_{obj} \ll h_{GLM}$ we get:

\begin{align}
  R^2 = 2r_{earth}^2\left(1-f(\theta_{obj}, \phi_{obj}, \phi_{GLM})\right) +2r_{earth}h_{GLM}\left(1-f(\theta_{obj}, \phi_{obj}, \phi_{GLM})\right) + h_{GLM}^2 - 2h_{GLM}h_{obj}f(\theta_{obj}, \phi_{obj}, \phi_{GLM})
\end{align}

\section{Useful databases}

>>>>

% This is the appendix, where we may insert detailed mathematical procedures or other data which may be useful but also will make the main
% text too large so this should be consulted only in case of necessity
\section{Bolide velocity components}
\label{app:velocity}
The USG database from JPL is a great source of information about the brightest bolides that have entered the atmosphere since 1988. However, the velocity components $(v_x, v_y, v_z)$ are displayed in a geocentric Earth-fixed reference frame explained as follows:

$v_x$ lies in the Earth's equatorial plane, parallel to the equator and points towards the prime meridian, $v_z$ is parallel to the earth's rotational axis and points towards the north celestial pole and $v_y$ completes the right-handed reference system. Figure  shows graphically these velocity components.

In the other hand, it might be useful and more intuitive to describe the velocity components in terms of the local reference system composed by $(v_{lon}, v_{lat}, v_h)$, where $v_{lon}$ is the velocity component parallel to the equatorial plane, positive towards east, $v_{lat}$ is parallel to the polar plane, positive to north, and $v_h$ is the radial component pointing towards the center of Earth. Figure \ref{fig:ref_velocity} sketches both reference systems simultaneously.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{../figures/meteor_velocity_references.pdf}
    \caption{Velocity components of bolide. The geocentric Earth fixed reference system is sketched in black and the local spherical reference system in blue. The components parallel to the equatorial plane are not shown since are perpendicular to the plane of the page.}
    \label{fig:ref_velocity}
\end{figure}

The transformation between both reference systems in the north-west hemisphere is given as follows:

\begin{align}
v_{lon} &= v_x \label{eq:vlon}\\
v_{lat} &= v_z\cos{L} - v_y\sin{L} \label{eq:vlat}\\
v_h     &= v_y\cos{L} + v_z\sin{L} \label{eq:vh}
\end{align}

Where $L$ is the bolide latitude. With this information, we can estimate the velocity tangential to the Earth's surface as $v_{tan} = \left((v_{lon})^2 + (v_{lat})^2\right)^{1/2}$ and finally the trajectory angle respect to the normal as $\tan\theta = \frac{v_{tan}}{v_h}$. Using equations (\ref{eq:vlon}) to (\ref{eq:vh}) we obtain the following:

\begin{align}
    v_{tan} &= \left((v_x)^2 + (v_z\cos{L}-v_y\sin{L})^2\right)^{1/2} \\
    \tan\theta &= \frac{\left((v_x)^2 + (v_z\cos{L}-v_y\sin{L})^2\right)^{1/2}}{v_y\cos{L} + v_z\sin{L}}  \label{eq:meteor-angle}
\end{align}

Substituting the corresponding data from table \ref{tab:Meteor-parameters} into equation (\ref{eq:meteor-angle}) we find that that the trajectory angle of the meteor respect to the normal is $\theta \approx 64.4^{\circ}$

\section{Detrending test signal}
\label{app:test-signal}
Figure \ref{fig:test-signal} shows the form of this test time series, described mathematically as follows \citep{Boris:2020}:
 \begin{align}
     I_R(t) &= A\cdot \exp\left[-0.5\left(\frac{t-t_m}{d_t}\right)^2\right]\cdot \sum^{i}_{n}\sin\left(\omega_i t\right) \label{eq:ref-signal}\\
     Trend(t) &= B\cdot \left|t - t_0\right|^3 \label{eq:trend}
 \end{align}

 \begin{figure*}
     \centering
     \includegraphics[width=\linewidth]{../figures/test_signal}
     \caption{Initial setup for testing detrending method. a) Reference signal given by equation (\ref{eq:ref-signal}). b) Superposition between a) and the trend given by (\ref{eq:trend})}.
     \label{fig:test-signal}
 \end{figure*}

 Where equation (\ref{eq:ref-signal}) is the signal we want to detrend and equation (\ref{eq:trend}) is the trend we want to remove. $A$ is the amplitude of the signal, set as $\SI{0.2}{TECU}$, $t_m$ is a parameter which determines the position of the envelope maximum, set as $\SI{250}{min}$, which is the half of the array length, $d_t$ is the half width of the envelope, set as $\SI{50}{min}$ and $\omega_i$ are the frequencies of three harmonics with periods of $\SI{20}{min}$, $\SI{40}{min}$ and $\SI{60}{min}$. In the other hand, $B$ is the amplitude of the trend, set as $\SI{3.84e-6}{TECU.min^{-1}}$ and $t_0=\SI{250}{min}$ determines position of the minimum of the trend.

 For the detrending we used a Savitsky-Golay filter of order 3, this is lowest order filter which at the same time avoids too much oscillations and allows to the fit (and its derivative) to be smooth. The other remaining parameter is the window size, i.e. the number of convolution coefficients neccesary for the regression, which must be an odd integer, greater than the order of the polynomial order and lower than the array size. Then, we estimated the detrended signal using all possible values for the window size and estimated the residuals as follows:

 \begin{align}
     residuals = \sum_{i}^N \left(d_i - s_i\right)^2
 \end{align}
 Where $N$ is the array size, $d_i$ is the value of the detrended signal at the time $t_i$ and $s_i$ is the reference signal at the same time. In figure  we show the behavior of residuals as a function of the window size relative to the array size. We noted that the residuals behavior is almost insensitive to the window size when its size is lower than than 60\% of the array size, when the errors start to grow exponentially, but going to more detail, we found that using a window size of about $1/4$ of the array length, such errors are minimal. The detrended test curve is shown in figure \ref{fig:detrended-test}, compared with the original signal, as well as the substraction of both curves. In an ideal situation it should be zero for all times, but even in this case the amplitude of the envelope is about the tenth part of the amplitude of the envelope of the signal.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{../figures/IRvsDetrended}
\label{fig:detrended-test}
\caption{Left: Original test signal (blue continuous curve) compared with the result of detrending the superposition of the test signal with the test trend (figure \ref{fig:test-signal}b), which is the orange dashed curve. Right: Substraction of both curves of left side. The amplitude of the envelope of this residuals qualitatively is about tenth percent of the amplitude of the original signal.}
\end{figure*}

\section{Coordinates transformation}
\label{app:azimuth}

In order to change coordinates form geocentric to local coordinates for the bolide position, we may assume the flat earth approximation since the GPS stations we used in our work are located near the place the bolide was detected, and the apparent curvature of Earth at the bolide height is almost zero. With this in mind, in figure \ref{fig:flat_earth_aproximation} we show the flat Earth approximation in order to transform the geocentric coordinates of the bolide (latitude and longitude) to their corresponding local coordinates azimuth and elevation. With the aid of this image, we derive the following equations:

\begin{align}
    \tan{Az} &= \frac{\Delta \lambda}{\Delta L} \\
    \tan{Ele} &= \frac{h}{r}
\end{align}
Where $r = R_E\left((\Delta L)^2 + (\delta\lambda)^2\right)^{1/2}$ and $(\delta L, \delta\lambda)$ are the separation in latitude and longitude of the bolide from the GPS station, respectively. Azimuth is measured from the north counterclockwise and elevation is measured from the Earth's horizon to zenith, in such way azimuth goes from $0^\circ$ to $360^\circ$ and elevation from $0^\circ$ to $90^\circ$.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{../figures/coord_transform_drawing.pdf}
    \caption{Sketch of flat Earth approximation for transforming the fireball position from geocentric coordinates to local coordinates. From this picture we can derive equations () to () to estimate the local azimuth and elevation of the bolide at the GPS station position. $\vec{R}$ is the vector pointing to the bolide position with origin in the GPS station. $r$ is the projection of $\vec{R}$ in the Earth's plane. $L_s$ and $L_B$ are the latitudes of the station and the bolide, respectively, while $\lambda_s$ and $\lambda_B$ their longitudes, $h$ is the bolide height and $R_E$ the radius of Earth.}
    \label{fig:flat_earth_aproximation}
\end{figure}
%To do such transformation we do the following process \citep{Kroger:1996}:

%The transformation from celestial coordinates to local coordinates are given by:

%\begin{align}
%    \sin(E) = \sin{\delta}\sin{L} + \cos{L}\cos{H}\cos\delta \label{eq:elevation}\\
%    \cos{A} = \frac{\sin{\delta}-\sin{E}\sin{L}}{\cos{E}\cos{L}} \label{eq:cos-A}
%\end{align}
%Where $E$ is the elevation of the object, $\delta$ is the declination, $H$ is the hour angle, $A = Az$ when $\sin{H} <0$, otherwise $Az = 360-A$ where $Az$ is the Azimuth. Finally $L$ is the latitude of the observer (i.e, the GPS station). 
%If there were an observer located in the same coordinates than the fragmentation of the bolide took place, then this hypothetical observer should have seen the bolide at zenith. The celestial coordinates of the bolide at this location are zero hour angle and declination equal to object reported latitude. The hour angle is relative to the time and place the object is being observed, and we need the hour angle of the object from the place is observed (the GPS station). To do this we need estimate the hour angle of the object from the station perspective, and the hour angle from the "zenith perspective":
%\begin{align}
%    H = LST_s - RA\label{eq:LST}\\
%    0 = LST_z - RA \label{eq:z-LST}
%\end{align}
%Where $LST_s$ is the Local Sidereal Time from the station perspective, and $LST_z$ is the same but from the "zenith perspective", and $RA$ is the right ascension, which is the same in both perspectives. We can also write the Local sidereal time in terms of the Greenwich local time as follows:%If we sum the last equations we find that:
%\begin{align}
%    LST = GST + Lon_{obs} -RA \label{eq:LST-GST}
%\end{align}

%Where $GST$ is the Greenwich Local time and $Lon_obs$ is the observer longitude. Substituting (\ref{eq:LST-GST}) into (\ref{eq:LST}) and (\ref{eq:z-LST}) and sum the resulting equations we find that:

%\begin{align}
%    H = Lon_s - Lon_z \label{eq:H-lon}
%\end{align}

%Where $Lon_s$ is the station longitude and $Lon_z$ is the longitude of the bolide. Substituting \ref{eq:H-lon} into (\ref{eq:elevation}) and (\ref{eq:cos-A}) we can find the Azimuth and elevation in terms of the local latitude and longitude and the object's reported latitude and longitude:

%\begin{align}
%    \sin(E) = \sin{\delta}\sin{L} + \cos{L}\cos(Lon_s - Lon_z) \label{eq:elevation_f}\\
%    \cos{A} = \frac{\sin{\delta}-\sin{E}\sin{L}}{\cos{E}\cos{L}} \label{eq:cos-A_f}
%\end{align}

%Examples of bolides trajectories against satellites trajectories are found in figures \ref{fig:boav-trajectory} - \ref{fig:ttsf-trajectory} for stations BOAV, CN00, CN04, CN05, KOUR and TTSF:

%\begin{figure}
%    \centering
%    \includegraphics[width=\linewidth]{../figures/azimuth-elevation-map-boav-polar}
%    \caption{Satellites trajectory from BOAV station perspective in colors in polar coordinates, where the azimuth is represented in the polar axis and the elevation in the radial axis, being the zenith at the center of the graph and the horizon at the edge. The estimated meteor trajectory is shown in a black dashed line and GLM data as magenta dots. Practically for all stations, the satellite with PRN 13 is the closest to the bolide trajectory. The dashed curves represent the propagation radius of TIDs after 1 hour (red) and two hours (blue) in the azimuth-elevation space assuming TIDs propagate at \SI{362}{km.s^{-1}}, the same as the reported high frequency TIDs from Chelyabinsk meteor \citep{Yang:2014}.}
%    \label{fig:boav-trajectory}
%\end{figure}

%\begin{figure}
%    \centering
%    \includegraphics[width=\linewidth]{../figures/azimuth-elevation-map-cn00-polar}
%    \caption{Same as figure \ref{fig:boav-trajectory} but using station CN00.}
%    \label{fig:cn00-trajectory}
%\end{figure}\begin{figure}
%    \centering
%    \includegraphics[width=\linewidth]{../figures/azimuth-elevation-map-cn04-polar}
%    \caption{Same as figure \ref{fig:boav-trajectory} but using station CN04.}
%    \label{fig:cn04-trajectory}
%\end{figure}\begin{figure}
%    \centering
%    \includegraphics[width=\linewidth]{../figures/azimuth-elevation-map-cn05-polar}
%    \caption{Same as figure \ref{fig:boav-trajectory} but using station CN05.}
%    \label{fig:cn05-trajectory}
%\end{figure}
%\begin{figure}
%    \centering
%    \includegraphics[width=\linewidth]{../figures/azimuth-elevation-map-kour-polar}
%    \caption{Same as figure \ref{fig:boav-trajectory} but using station KOUR.}
%    \label{fig:kour-trajectory}
%\end{figure}
%\begin{figure}
%    \centering
%    \includegraphics[width=\linewidth]{../figures/azimuth-elevation-map-ttsf-polar}
%    \caption{Same as figure \ref{fig:boav-trajectory} but using station TTSF.}
%    \label{fig:ttsf-trajectory}
%\end{figure}

\section{UNAVCO Acknowledgements and stations list}

In table  we will enlist the stations from we collected data for our work. This material is based on services provided by the GAGE Facility, operated by UNAVCO, Inc., with support from the National Science Foundation and the National Aeronautics and Space Administration under NSF Cooperative Agreement EAR-1724794, we are deeply grateful with all the people that was involved and whose work we are citing here.

\clearpage
\onecolumn
\footnotesize
\begin{landscape}
\begin{table*}
    \centering
       \caption{List of GPS stations used for this work.}
      \label{tab:table-stations}
    \begin{tabular}{lllp{8cm}}

    %  \endfirsthead
    %  \endhead
    \toprule
    Station name & Latitude (deg) & Longitude (deg) & Citation \\
    AIRS & 16.74 & -62.21 & \url{https://doi.org/10.7283/T53B5XGJ}\\
    BARA & 18.21 & -71.09 & None available                       \\
    BOAV & 2.85  & -60.70 & None available \\
    CN00 & 17.67 & -61.79 & \url{https://doi.org/10.7283/T5FN14GQ} \\
    CN04 & 14.02 & -60.97 &  \url{https://doi.org/10.7283/T5BP0124} \\
    CN05 & 18.56 & -68.35 & \url{https://doi.org/10.7283/T5VQ30ZH} \\
    CN19 & 12.61 & -70.04 & \url{https://doi.org/10.7283/T5HD7SZB}\\
    CN27 & 19.67 & -69.93 & \url{https://doi.org/10.7283/T5JD4V2P} \\
    CN40 & 12.18 & -68.96 & \url{https://doi.org/10.7283/T5BV7DWT} \\
    CRLR & 18.41 & -68.93 & \url{https://doi.org/10.7283/T5FN14JM} \\
    CRSE & 18.76 & -69.04 & None available \\
    GERD & 16.80 & -62.19 & \url{https://doi.org/10.7283/T5TT4PBT} \\
    GRE1 & 12.22 & -61.64 & \url{https://doi.org/10.7283/T5BC3WZ5} \\
    JME2 & 18.23 & -72.54 &  \url{https://doi.org/10.7283/T5KW5D38}\\
    KOUG & 5.10  & -52.64 & None available \\
    KOUR & 5.25  & -52.81 & None available \\
    LVEG & 19.22 & -70.53 & \url{https://doi.org/10.7283/T5CZ35GC}\\
    NWBL & 16.82 & -62.20 & \url{https://doi.org/10.7283/T5ZK5F13} \\
    OLVN & 16.75 & -62.23 & \url{https://doi.org/10.7283/T5Q23XMD} \\
    RCHY & 16.70 & -62.15 & \url{https://doi.org/10.7283/T5707ZSJ} \\
    RDAZ & 18.45 & -70.72 & None available \\
    RDF2 & 19.45 & -70.68 & None available \\
    RDHI & 18.60 & -68.72 & None available \\
    RDLT & 19.31 & -69.55 & \url{https://doi.org/10.7283/T5J101GT} \\
    RDMA & 19.54 & -71.08 & \url{https://doi.org/10.7283/T50863NM} \\
    RDMC & 19.85 & -71.64 & None available \\
    RDMS & 18.98 & -69.04 & \url{https://doi.org/10.7283/T5DV1HPQ} \\
    RDNE & 18.50 & -71.42 & None available \\
    RDON & 16.93 & -62.35 & \url{https://doi.org/10.7283/T5W37TFB} \\
    RDSD & 18.46 & -69.91 & \url{https://doi.org/10.7283/T5CZ3594} \\
    RDSF & 19.29 & -70.25 & None available \\
    RDSJ & 18.82 & -71.23 & \url{https://doi.org/10.7283/T59W0CTW} \\
    SPED & 18.46 & -69.31 & \url{https://doi.org/10.7283/T5HQ3X75} \\
    SROD & 19.48 & -71.34 & \url{https://doi.org/10.7283/T5862DSD} \\
    TGDR & 18.21 & -71.10 & \url{https://doi.org/10.7283/T5222S3R} \\
    TRNT & 16.76 & -62.16 & \url{https://doi.org/10.7283/T5K935W2} \\
    TTSF & 10.28 & -61.47 & \url{https://doi.org/10.7283/T5JQ0ZCJ} \\
    TTUW & 10.64 & -61.40 & \url{https://doi.org/10.7283/T5TQ5ZTR} \\
    \bottomrule
\end{tabular}     
\end{table*}
  \end{landscape}
  \clearpage
  \twocolumn

### 2022-11-29 paper/discussion.tex
Update ELSEVIER paper (which compiles) to overleaf at 28-November-2022

We collected data for a sample of meteors which were detected around Mexican territory, or with some exceptions, meteoroids that released enough energy to have a good chance to get a vTEC disturbance with nearby GPS stations. We used data from the GLM, where the total energy was estimated from light curves, and USG sensors, where energy data is available. We used the few events which are present in both samples to re-calibrate the energy obtained for the GLM sample since in this few cases the energy estimated with the GLM data is systematically lower than the energy reported by USG sensors. Using the resulting energies, we estimated the bolides mass using a Fragment Cloud Model assuming a certain velocity range.

We have collected GPS data of nearby stations where the event was detected at the date each event occured, and one day before as comparison. We obtained the the vTEC series and we detrended them to remove the effects of Earth rotation and solar activity and make the wave-like features more evident. With these data, we used a wavelet transform using a Morlet wavelet to discover the frequencies at which vTEC perturbations may be produced, how intense these perturbations are and the time interval of these perturbations. If two different stations detect perturbations at the same frequencies that means that a vTEC perturbation was detected with such stations. We have observed that the distribution of wavelet coherent signal show us that the propagation of ionospheric disturbances have a time intervals of occurrence of a two hours from the spectrograms.

>>>>

We collected data for a sample of GPS station from UNAVCO network around the Caribbean bolide was detected in the same day and next day. We used data from the GLM, where the total energy was estimated from light curves, and USG sensors, where energy data is available to spot the bolide position and estimate the trajectory at the fragmentation instant. 
For GPS data, we adapted the method from \citet{Pradipta:2015} for detrending all the resultant TEC curves to remove the effects of Earth rotation and solar activity and make the wave-like features more evident. In order to discriminate which stations made a detection from the ones who didn't, we compared the detrended time series with a Morlet wavelet. We discarded other sources of GWs such like slow solar wind reached by fast wind, fluctuations in solar X-rays, coronal mass ejections and solar terminator effects. We found TIDs features in sTEC series from stations CN04, CN40, GRE1, TTSF, TTUW, BOAV, KOUG and KOUR stations with at least one satellite.

### 2022-11-29 paper/introduction.tex
Update ELSEVIER paper (which compiles) to overleaf at 28-November-2022

The Earth's magnetic field represents a final obstacle to the Solar Wind (SW) flux. When decelerated and defected by a non colisional shock wave in the flux direction, generates a cavity known as magnetosphere \citep{Blanco-Cano:2004}. Since the Earth is embedded in this SW flux, is known that under adequate physical conditions (e.g magnetic reconnection) may exist some coupling between the magnetosphere and the Earth's ionosphere \citep{Zolesi:2014, Cnossen:2014}.

The Sun plays an important role in the physical processes that occur in the terrestrial magnetosphere-ionosphere system. When the SW interacts with the Earth's magnetosphere, particles may permeate the internal region via magnetic reconnection and penetrate to polar zones and generate boreal or austral auroras thus altering the system \citep{Vazquez:2016, Oka:2011}. By the other hand, the Extreme Ultraviolet Radiation (EUV) and X-rays coming from the Sun may interact with the neutral atmoshere via photoionization \citep{Vlasov:2010}. However, in both cases the final result is that the ionosphere's free electrons population is altered.

Some Ionospheric Perturbations (IP) become relevant due to their spatial and temporal scale in the Space Weather scenario. At intermediate latitudes, the most common in the ionosphere are known as Traveling Ionospheric Disturbances (TIDs). Typically they divide into two groups: a) large scale TIDs, associated with geomagnetic storms with sizes of \SI{\sim 2000}{km}, periods of \SI{\sim 1}{h} and velocities of \SI{\sim 700}{km.s^{-1}}, and b) Medium-scale TIDs, which are not fully associated with geomagnetic storms, present sizes of \SI{\sim 100}{km}, periods from 10 minutes to 1 hour and velocities between \SI{50}{km.s^{-1}} and \SI{1e2}{km.s^{-1}} \citep{Helmboldt:2012}. Diverse methods have benn used to study TIDs, such as incoherent dispersion radars, high frequency Doppler emmisors, data from Global Positioning System (GPS) stations or even radiotelescopes like the VLA or the Mexican Array Radio Telescope (MEXART) \citep{Chilcote:2015, Rodriguez:2014}.

On the other side, the Earth's ionosphere may be affected or modified by other proccesses, particularly there are studies that show how the Vetical Total Electron Content (vTEC) due to shock waves generated for rockets launched to space \citep{Lin:2014}. Similar proccesses modify the Earth's ionosphere due to objects entering the athmosphere from space, such as meteoroids like the one which fell on Chelyabinsk at 2013 \citep{Yang:2014}. Previously, the ionospheric perturbations produced by this object were studied using two independent methods: a) detecting vTEC pertubations using GPS station near the impact location. And b) a wavelets analysis for detection of ...

In 2020 a meteoroid passed in mexican territory through mexican territory, which also was studied \citep{Sergeeva:2020}. The meteoroid was recorded with outdoor cameras in different locations. The trajectory could be estimated, as well as other physical parameters.

In this work we will show a similar analysis for a sample of meteoroids detected in Mexican territory by different methods. The first subsample consists in objects detected by the Geostationary Lightning Mapper (GLM) whose sizes are estimated between a few decimeters to meters in diameter \citep{GOODMAN:2013, Jenniskens:2018, Rumpf:2019}. The second subsample will consist in objects detected by ocular witnesses from the American Meteor Society and as comparisson we will include the morelian meteoroid reported in \citet{Sergeeva:2020} and the Chelyabinsk event \citet{Yang:2014}. The paper is arranged in the following way: \S \ref{sec:methodology} describes the samples of meteoroids as well of the properties that can be obtained from direct observations. Also describes the GPS data corresponding to the dates and locations where each object was located. \S \ref{sec:bolides} shows physical parameters of meteoroids obtained from the observed heights and energies. Finally, section \S \ref{sec:vTEC-maps} shows the vTEC maps and scintillation indices obtained from GPS observations.

>>>>

Bolides impact the Earth every day, from the smallest ($d<\SI{0.5}{m}$), which are destroyed when entering the atmosphere to the biggest and most rare events, where an asteroid with a diameter $d>\SI{100}{m}$ is almost unaffected by the atmosphere and is technically unable to slow them down. If their diameter is greater than \SI{1}{km} the impact can be considered a global catastrophe \citep{Dudorov:2020}. Some famous modern cases are the Tunguska event 1n 1908 \citep{Wheeler:2019} and the Chelyabinsk event \citep{Yang:2014} in 2013.

Such events have been recently recorded in certain databases, such as the Center for Near Earth Objects Studies (CNEOS), Jet Propulsion Laboratory (JPL) and the Geostationary Lightning Mapper (GLM), publicly available at \url{https://cneos.jpl.nasa.gov/fireballs/} and \url{https://neo-bolide.ndc.nasa.gov/#/}, respectively.

We picked our interest in one certain event, that we nicknamed ``the Caribbean Meteoroid'', because this meteoroid is the most energetic which appear in both databases and thus is an interesting case for studying the potential Traveling Ionospheric Disturbances (TIDs) that could be produced. To do so, we collected RINEX data from nearby GPS stations from the UNAVCO network (also publicly available) and developed a method to detrend the resultant TEC series and estimate the TIDs propagation speed ... (The introduction also needs to be polished after getting the main results)

### 2022-11-29 paper/meteor-properties.tex
Update ELSEVIER paper (which compiles) to overleaf at 28-November-2022

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{../figures/Energy_vs_Mass-fireballs_en}
    \caption{Bolides mass derived from Fragment Cloud Model. We used three assumptions about the bolide velocity: Minimum velocity $(\sim \SI{15}{km.s^{-1}})$, Medium velocity $(\sim \SI{16}{km.s^{-1}})$ and maximum velocity $(\sim \SI{18}{km.s^{-1}})$. This values are based on the available data of USG sample (see table \ref{tab:table-meteors-2})}
    \label{fig:energy-mass}
\end{figure}

Once the estimated the bolides energy, we are capable to estimate other bolides parameters, such as the mass. For that we used a Fragment Cloud Model \citep{Wheeler:2017}. To obtain the bolide mass, we must have previous knowledge of its velocity. Based on the few events of the USG sample where we could extract the velocity (see table \ref{tab:table-meteors-2}), we tried three models of bolides mass as a function of its releasing energy: the low velocity model, which assumes a velocity of $\sim \SI{15}{km.s{^-1}}$, medium velocty $v \sim \SI{16}{km.s{^-1}}$ and high velocity $v \sim \SI{18}{km.s{^-1}}$. The resultant masses can be seen in figure \ref{fig:energy-mass}. For the rest of the sample we need to assume that the bolides velocity is within this range or close enough in order to get a reasonable estimation of its mass.

>>>>

\begin{table}
    \centering
     \caption{Caribbean meteor possible physical properties obtained with  algorithm  , assuming a density of \SI{3000}{kg.m^{-3}}}
    \begin{tabular}{lr}\toprule
       Initial Velocity  &  \SI{19.9}{km.s^{-1}}\\
        Initial mass & \SI{340}{kg} \\
        Entry angle (deg) & 64 \\
        \bottomrule
    \end{tabular}
    \label{tab:meteor-properties}
\end{table}

Using the estimation of the meteor energy and velocity from table \ref{tab:Meteor-parameters}, the trajectory angle estimated in appendix \ref{app:velocity} from the velocity components of $64.4^\circ$ and assuming the meteor has a density of \SI{3000}{kg.m^-3}, consistent with the composition of \textit{condrites} (we must assure or edit this last statement), we estimated the initial velocity, initial mass and entry angle using (add citation and more text and once got the references) ...%we are capable to estimate other bolides parameters, such as the mass. For that we used a Fragment Cloud Model \citep{Wheeler:2017}. To obtain the bolide mass, we must have previous knowledge of its velocity. Based on the few events of the USG sample where we could extract the velocity (see table \ref{tab:table-meteors-2}), we tried three models of bolides mass as a function of its releasing energy: the low velocity model, which assumes a velocity of $\sim \SI{15}{km.s{^-1}}$, medium velocty $v \sim \SI{16}{km.s{^-1}}$ and high velocity $v \sim \SI{18}{km.s{^-1}}$. The resultant masses can be seen in figure \ref{fig:energy-mass}. For the rest of the sample we need to assume that the bolides velocity is within this range or close enough in order to get a reasonable estimation of its mass.
